What We Eval in the Shadows

### 2021-04-16 paper/main.tex
Updates in 6

\subsection{Base}

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute}, and \c{eval} are the tools given to
developers to this end. This API is slightly more complex than just
passing a string, it is conceivable that this may discourage some
casual users. \Eval is also being used to reduce boilerplate code and
provide convenience features for programmers. We give examples for the minimized expressions of Section~\ref{sec:minimized} and then dig into high-level \eval design patterns.

The examples are based on code from CRAN or Kaggle, simplified to fit the limited space of a scientific paper. For each example, we indicate the package and the function where it is located.

@@

We illustrate the minimized expressions of Section~\ref{sec:minimized} on representative examples.

% model.frame is in the next subsection, in the high-level patterns

\subsubsection{Variable lookup and values.}  We also include here lookup to slots in a list or data frame.
This corresponds to minimized expressions \c{X}, \c{V}, and  \c{\$}. Variable lookup leads to a large diversity of environments, many being package environments or package environments wrapped in \c{new}.

It is the predominant usage of \eval in Base, through function \c{match.args} (with evaluates on  minimized expression \c{V}). In CRAN, it correlates with more exotic environments passed to \eval. 
The dominant site in numbers of calls in CRAN, is located in \c{ggproto} in the custom object system of \emph{ggplot2}. It evaluates a variable \c{`\_inherit`}, which represents the parent of the class, in the definition environment of the class. \eval is used here to account for parents scoped in a namespace with \c{::}, which is actually a function call in R.

Package \emph{statnet.common} handles formulas, R object that associates an expression of the form \c{x~\textasciitilde~expr} and a carry-on environment where to evaluate the formula. It evaluates the left-hand side of the formula, which is always a variable, in function \c{eval\_lhs.formula}, where \c{object} is the formula and \c{environment} returns the formula associated to the object:

\begin{lstlisting}
eval(object[[2L]],envir=environment(object))
\end{lstlisting}

\subsubsection{Assigments.} It corresponds to minimized expression \c{<-} and groups together \c{<-}, \c{<<-} and \c{assign}. Package \emph{plyr} has a function combine data frames by row, filling in missing columns. During the allocation of the missing columns, an assignment call is built with \c{assignment <- quote(column[rows] <<- what)} and later evaluated on demand in a setter function, with \c{eval(assigmment)}. Package \emph{overture} is used to run Markov Chain Monte Carlo. It runs an R expression repeatedly and saves any assignments that happen, as samples of the Markov chains. 
The expression is captured with \c{substitute} and a new environment \c{env} wrapping the parent of the caller is created, in which all assignments will happen. Regularly, the content of \c{env} is inspected to sample the assigned variables.
\begin{lstlisting}
function(expr, overwrite=over.write) {
	expr_q <- substitute(expr)
	env <- new.env(parent=parent.frame(1))
	samps <- list()
	RunMcmc(samps, 1, expr_q, env, n.save, backing.path, thin, exclude, overwrite)
}

RunMcmc <- function... {
	 eval(expr_q, envir=env)
\end{lstlisting}

\subsubsection{Function definition.} This corresponds to minimized expression \c{FUN}. Package \emph{Rcpp} is used to easily interface C++ and R and it uses \eval to bind C++ functions and classes to R objects.
For instance, \c{.makeCppMethods} create a R function for each C++ method named \c{what} of a given class. The use of \substitute to inject the name of the method in a language expression entails the use of \eval.
\begin{lstlisting}
methods[[what]] <- eval(substitute(
	function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))),
	env)
\end{lstlisting}

\mypara{Block.} In package \emph{vglm}, \eval is used to inject a pre-processing code at the beginning of a function implementing a statistical model: 
\begin{lstlisting}
if (length(slot(family, "first"))) 
	eval(slot(family, "first"))
\end{lstlisting}

Package \emph{withr} provides facilities to encapsulate side-effecting functions. The function \c{defer} is used to attach an expression to be evaluated when exiting any frame in the stack. It can be passed arbitrary code, and so blocks.

\subsubsection{Simple function calls.} This corresponds to minimized expression \c{F()}, \ie calls without nested functions and without variables (but possibly named arguments). For instance, package \emph{foreach} provides \c{foreach}-like loops. It makes it possible to initialize the iterator variables, with code \c{init}, which is equal \c{quote(list())} by default. \c{init} is then used to initialize the \c{foreach} accumulator \c{state\$accum <- eval(combineInfo\$init, envir=extra, enclos=obj\$evalenv)}.

A peculiar use of \eval is to bypass some checks performed to accept code submitted to CRAN. For example, bindings can be \emph{locked} in an environment, \ie forbidding its value to be modified. The binding can be unlocked though by calling \c{unlockBinding} but packages in CRAN are forbidden to use that function.  Package \emph{data.table} bypasses this restriction by building a call piece by piece, to unlock binding \c{.N}. The call is then evaluated. Now that \c{.N} is unlocked, it can be assigned to. Instead of using \eval to assign in the parent frame, they rather use the more specialized \c{assign} function.

\begin{lstlisting}
	if (locked.N) eval(call("unlockBinding", ".N", parent.frame()))  
	assign(".N", nrow(x), envir=parent.frame(), inherits=FALSE)
\end{lstlisting}

\subsubsection{Complex function calls.} It corresponds to minimized expressions \c{F(X)} and \c{F(F(X))}, \ie function calls with variables, possibly with nested function calls. Package \emph{huxtable} is used to create custom tables and export them to various formats. To create functions to customize each of the borders of the table, it injects the indices \c{i} and \c{j} of the cells at the border and then evaluate them:
\begin{lstlisting}
xxx_border_huxtable <- function(lr_tb, i, j, border_prop) {
	function (ht) {
		hickness <- mlist$thickness[eval(i), eval(j), drop = FALSE]
	}
}
left_border.huxtable <- xxx_border_huxtable("lr_borders",
i = quote(seq_len(nrow(ht))), j = quote(-last_col))
\end{lstlisting}

>>>>

\subsection{Base}\label{sec:usage-base}

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute} and \c{eval} are the tools given to developers to
this end.
%This API is slightly more complex than just passing a string, it is
%conceivable that this may discourage some casual users. \Eval is also being
%used to reduce boilerplate code and provide convenience features for
%programmers. 
In this section we focus on the \emph{why}, giving real-world examples of the
\eval usage in the wild. We start the discussion with the examples for the
minimized expressions from Section~\ref{sec:minimized}. They we look into
high-level \eval design patterns. All of the examples are based on code from
our corpus often simplified to fit the space and increase clarity. For each
example, we indicate the package and the function where it is located.

@@

In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of expressions passed to \eval.
Here we illustrate them on concrete instances.

\begin{compactitem}[---]

  \item \emph{Variable lookup and values and indexing \c{(X, V, \$)}}
    constitute one of the most common use case for \eval. Let's consider the
    \c{subset} function from R base library which allows, among others, to
    select variables in a data frame. For example, given a data frame \c{df}
    with three columns \c{A,B,C}, \c{subset(df, sel=A)} returns a subset of
    \c{df} with just the \c{A} column. The same would happen with \c{subset(df,
    sel=1)}. However, thanks to \eval it can also be also called with
    \c{subset(df, sel=c(1,B,x))} returning a subset of the first column, column
    \c{B} and whatever index the variable \c{x} holds. It is implemented using:
    \lstinline|eval(substitute(sel), nl, parent.frame())| The first argument
    returns the expression passed to \c{sel}, the second is the data frame in
    which this expression will be evaluated and finally the last indicates
    which environment should be used to look up bindings not found in \c{nl}.

    Another example is the most often called \eval site in CRAN:
    \lstinline|eval(`_inherit`, env)| defined in the \c{find\_super} function
    in \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
    find the parent class. In the vast majority of cases the variable is symbol
    and so the \eval seems superfluous. The \eval is however necessary so one
    can specify parent classes using \c{pkg\_name::class\_name} as the \c{::}
    operator is a function call in R.

  \item \emph{Assignments \c{(<-)}} appears in the cases where a developer wants
    to create or update a biding in a specific environment where either the
    name of the value come from some expression. Most cases come from trivial
    code generation where the assign expression is assembled using \c{parse} or
    \c{substitute}, often in a loop: \lstinline|eval(parse(text=paste(capture.output(default_args[[i]]), "<-", i)))|.

    There are also more sophisticated cases. The \c{plyr::allocate\_column}
    function uses \eval for a deferred assignment to fill missing values in a
    column. Package \c{overture} repeatedly evaluates R expressions saving any
    assignments that happen in it, as samples of Markov chains.

  \item \emph{Function definition \c{(FUN)}} is mostly used in conjunction with
    code generation where new functions are synthesized using either \c{parse}
    or \c{substitute}. Some FFI frameworks use this to automatically generate R
    binding to native code. For example, \c{Rcpp}, which bridges R with C++,
    uses \eval to generate R function for C++ methods:\\
    \lstinline|eval(substitute(function(...) .CppObj$M(...), list(M=...)), env)|.

  \item \emph{Block \c{(BLOCK)}} can appear anywhere, but it is almost always
    the case that the block is only passed to \eval rather than being directly
    part of in as in \c{eval({...})}. Essentially, the block denotes a fragment
    of a program to be evaluated in a special way. There is a number of use
    cases: unit testing frameworks (\eg \c{testthat}, \c{testit}), code
    benchmarking (\eg \c{rbenchmark}, \c{microbenchmark}), running code in
    parallel (\eg \c{foreach}, \c{doParallel}), or deferring code execution
    (\eg \c{withr}).

  \item \emph{Simple function calls \c{(F())}}

    A peculiar use of \eval is to bypass some checks performed by the \c{R CMD
      CHECK} tool allowing one to submit a package to CRAN that would otherwise
      be refused. For example, the tool checks that packages do not call
      \c{unlockBinding} function.\footnote{It allows one to open and modify
      loaded package name space.} Packages such as \c{data.table} circumvent
      such restrictions by using \eval as the check is only performed
      statically.

  \item \emph{Complex function calls \c{(F(X), F(F(X)))}}
